%!TEX TS-program = xelatex
%!TEX encoding = UTF-8 Unicode

\documentclass[12pt,a4paper]{article}
\usepackage[margin=1.5in]{geometry}
\usepackage[parfill]{parskip}
\usepackage[superscript,biblabel]{cite}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}

\usepackage[unicode]{hyperref}
\hypersetup{pdfauthor={Kenrick Turner}, pdftitle={A pre-deployment medical plan for Rothera Research Station, Antarctica}, pdfsubject={Antarctic medical risk assessment},pdfkeywords={antarctica, risk assessment, medical}}

% Need this to rotate table in appendix
\usepackage{pdflscape}

% Pretty table stuff
\usepackage{tabu}
\usepackage{longtable}
\usepackage{booktabs}

\usepackage{fontspec,xltxtra,xunicode}
\defaultfontfeatures{Mapping=tex-text}
\setmainfont[Mapping=tex-text]{Helvetica Neue}
\setmonofont[Scale=0.9]{Source Code Pro}
\usepackage{microtype}

% Fancy bolded numbers for enumerate environment
\newenvironment{boldenumerate}
    {\begin{enumerate}\renewcommand\labelenumi{\textbf\theenumi}}
    {\end{enumerate}}

\title{A pre-deployment medical plan for Rothera Research Station, Antarctica}
\author{REM711: Pre-expedition Medical Plan}
\date{Student: 10463823 \\ Word Count: 4925}

\linespread{1.3}

\begin{document}

\maketitle

\tableofcontents

\pagebreak

\section{Introduction}

\begin{quote}
Proper Planning and Preparation Prevents Poor Performance. \\
\em British Army adage
\end{quote}

In October 2013, the British Antarctic Survey (BAS) will begin its summer operations and resupply of Rothera Research Station, involving the transport of over a hundred personnel to Antarctica and a steady ramp up in scientific activities for the Austral summer. Personnel deploy to the continent for a variety of durations: from the legendary ``long-weekend in Rothera'' to an 18 month overwintering tour.

The Antarctic environment is frequently hostile and presents many hazards alien to the UK. Moreover, the ability to manage these hazards is severely constrained due to the limited infrastructure and personnel available. These limitations may require unorthodox or unusual solutions to be employed when devising a pragmatic medical plan.

This medical plan documents the risk assessment that has been undertaken to identify the range of risks presented by Antarctic operations. Whilst addressing general considerations, hazards identified as taking the greatest importance will be discussed comprehensively, including strategies for avoidance and control of the encountered risks, and a response plan in the event of harm occurring.

\section{Situation Briefing}

\subsection{Environment}

Rothera research station was established in 1975 and is located at 67°34'S, 68°08'W on Adelaide Island on the Antarctic peninsula. It lies 1,900km south of the Falkland Isles and 1,500km south-east of Cape Horn. The terrain surrounding Rothera is mountainous with peaks in exceeding 2,500m and is also heavily glaciated. Temperatures range from around 0 -- 5ºC during summer to between -5ºC and -20ºC in winter, with gales and high winds being commonplace. Lying within the Antarctic circle, Rothera experiences several weeks without sunlight during the winter.

The station comprises of several low buildings, including accommodation facilities, workshops, laboratories, and a control tower. Air operations are facilitated by a hangar, fuel depot, and 900m long runway and apron in close proximity to accommodation buildings. As a coastal station, Rothera also has a wharf permitting resupply by ship.

Rothera is resupplied by both sea and air. Cargo is predominantly transported by sea on one of two calls per season by BAS-operated vessels, whilst personnel movements usually occur by air on Dash-7 or Twin Otter aircraft.

\subsection{Population}

As the largest research station operated by BAS, the summer population approaches 130 staff, whilst for the 2013-14 season approximately 20 people are expected to overwinter. Personnel ages range from early twenties through to early sixties. There is a significant gender skew with a 4:1 male:female ratio amongst overwintering staff. All personnel are prescreened prior to deployment which includes both medical and dental assessment.

\subsection{Activities}

 Scientific research consists mainly of marine biology, geology, glaciology and meteorology, with the majority of the work conducted during the Austral summer.

In addition to conducting research on and around station, Rothera also serves as a logistics hub for UK science parties and partner nations operating in the deep field, with personnel transiting Rothera either by air or by overland travel on snowmobile or snowcat. Deep field parties are geographically and temporally remote -- frequently hundreds kilometres and several hours flight time away.

In addition to local and deep field operations, two forward operating bases fall under Rothera's medical cover. Fossil Bluff (71°20'S, 68°17'W) and Sky-Blu (74°51'S, 71°34'W) are two logistics facilities around 90 minutes and 3 hours flight time from Rothera respectively. As summer-only stations, their role is to support deep field operations and have only limited personnel on site living in pyramid tents. Communication with Rothera is conducted via radio, and there are comprehensive medical kits at both sites.

\subsection{Medical Resources}

\subsubsection{Primary Medical Care}

Medical cover on station is provided by a single Medical Officer (MO), assisted by several Advanced First Aiders -- trained in part by the MO -- and all personnel have basic First Aid training prior to deployment. During the Austral summer, there is an overlap period of several months between incoming and outgoing MOs whereby two doctors may be available on site.

Further medical support is rendered by a comprehensive telemedicine service by the British Antarctic Survey Medical Unit (BASMU), based in Derriford Hospital, Plymouth, UK. BASMU also trains MOs prior to deployment, which fosters a close working relationship between MOs and clinicians who will potentially be providing telemedicine support.

The medical bay at Rothera is physically small but well-provisioned, including plain film radiology, full anaesthetic capability, and \emph{in extremis}, general surgical capability.

\subsubsection{Secondary Medical Care}

The closest secondary medical care facility is Clinica Magallanes\cite{Anonymous:TgM8hVWD} in Punta Arenas, Chile, and is 1,600km and 4.5 hours flight time away by Dash-7 aircraft. It is a private tertiary referral centre for the region, providing all major medical and surgical specialities, including neurosurgery. It has both CT and MRI imaging available onsite.

An alternative is Kind Edward VII Memorial Hospital in Stanley, Falkland Islands. It offers 17 acute beds, with an emergency department, dental services, a single theatre, and 2 ITU beds.\cite{Anonymous:qFBtPcmx} Whilst it can provide plain film radiology it has no CT scanner. It is a proven evacuation route, having previously received medivacs from Rothera, but it is further away than Punta (1,900km), incurring an additional hour's flight time (5.5 hours). Moreover, Punta offers a more comprehensive range of facilities so should be the preferred evacuation route.

Medical evacuation normally takes place by air, using Dash-7 aircraft which have a pressurised cabin. However, aircraft are not normally stationed at Rothera during winter, and BASMU maintains a policy of not undertaking evacuations during this period due to the risks presented by inclement weather. An alternative method is by ship, and casualties have previously been evacuated under exceptional circumstances during winter via this route. It should be stressed that evacuation may be delayed for a considerable period in both summer and winter owing to poor conditions.

\section{Risk Assessment}

Risk is an integral part of every challenging human endeavour. Indeed, it often contributes to the appeal of undertaking exigent activity itself. When devising a medical plan, the process of logically assessing likely hazards and their risk harm is a valuable planning tool. This plan employs a qualitative risk assessment model, and defines a hazard as any agent that may cause harm, whereas risk may be understood as a qualitative description of the composite of the severity of the harm and its likelihood in occurring.\cite{TheNationalPatientSafetyAgency:2007ud}

The Health and Safety Executive (HSE) describes five steps\cite{HSE:2012tp} in the process of risk assessment:

\begin{enumerate}
    \item Identify the hazards
    \item Decide who might be harmed and how
    \item Evaluate the risks and decide on precautions
    \item Record the findings and implement them
    \item Review the assessment and update as necessary
\end{enumerate}

Using this model, a brief qualitative risk assessment for Rothera station is undertaken in Appendix~\ref{sec:risk}.

When a risk is identified, four techniques may be employed to manage it:

\begin{description}
    \item[Avoidance] aims to reduce the \emph{probability} of risk i.e. reducing the likelihood of a harm occurring -- preferably to zero. Whilst effective, taken to an extreme it entails not performing the activity that exposes people to the hazard which is often impractical.
    \item[Mitigation] aims to reduce the \emph{severity} of a risk i.e. reducing the harm created when a risk occurs. It is useful in situations where risk is sizeable and cannot be easily avoided.
    \item[Transference] shifts or shares the harm of a risk when it occurs to a third party. Transferring risk is most appropriate for situations where the harm of a risk can be clearly measured and fully addressed e.g. duplicating mission-critical equipment in case of its loss.
    \item[Acceptance] is the strategy of actively deciding that one will accept the harm of a risk should it occur and planning appropriately to manage the consequences. Clearly the potential gains should heavily outweigh the harm, so acceptance is best reserved for when the severity of harm is significantly smaller than the cost of avoiding, mitigating, or transferring the risk.
\end{description}

Whilst a fully comprehensive risk assessment for Rothera is beyond the scope of this medical plan, three risks have been identified that merit further discussion below.

\section{Challenges as the Lone Practitioner}

\begin{quote}
Physician, heal thyself. \\
\em Luke 4:23
\end{quote}

An inherent risk to Rothera is that for a significant proportion of the year there is a single medical provider. As a single point of failure, if the MO is unwell or injured, there is no other doctor to treat him. This risk may be mitigated by training advanced first aiders to a level where they can treat common conditions, and telemedicine support from BASMU transfers some of this risk. Furthermore, being aware of the risk, the MO may elect to avoid very high risk activities.

The doctor-patient relationship may also be altered when treating a colleague or friend, as professional judgement may be clouded by interpersonal issues. If there is a breakdown in friendship or working relationship, it may be difficult for an MO to act impartially. The facility to discuss issues with the BASMU telemedicine service provides a valuable sounding board for the MO in this context.

Moreover, the MO's actions may be scrutinised with disproportionate significance by other personnel, so the MO should be assiduously impartial to maintain their approachability and neutrality.

There is also a risk of developing a romantic relationship with patients in an isolated and confined environment, although this is discouraged. The GMC's guidance is unequivocal: ``You must not pursue a sexual or improper emotional relationship with a current patient.''\cite{Anonymous:2013wj,Anonymous:2013ve}

Confidentiality may also present a challenge, as the MO has a duty of care to not only the patient but to the station as a whole. There are therefore circumstances where the MO may need to breach confidentiality to the Base Commander to ensure the safety of others. Additionally, if a patient requires secondary medical care, logistical arrangements would also require limited disclosure of confidential information. Whenever the MO is disclosing confidential information, he should seek permission from the patient, explaining why and what is being disclosed.\cite{Anonymous:2013ta}

Tightly coupled to this issue is that the majority of the MO's work occurs in private, shielded from the rest of the station. Not only may this give rise to the impression of idleness by the MO, this situation also means the MO does not routinely have colleagues to vent or casually discuss confidences with -- often a key coping mechanism. Encouraging regular telephone contact with BASMU and other MOs on different stations is one mitigation strategy.

\section{Specific Risks}

\subsection{Accidental hypothermia}

\subsubsection{Rationale}

Accidental hypothermia is a potentially life-threatening involuntary decrease of ≥2ºC in normal core body temperature. Whilst its incidence is reportedly low in Antarctica\cite{BritishAntarcticSurveyMedicalUnit:2013vj} (probably due to robust education and provision of appropriate clothing), mild cases may go unrecognised.\cite{Brown:2012ja} Further, a brief risk assessment (Appendix~\ref{sec:risk}) demonstrates that the cold environment is a hazard common to many regular activities undertaken at Rothera, with hypothermia being one of the major associated harms.

\subsubsection{Risk assessment}

The principle hazard for hypothermia is cold exposure. Hypothermia occurs when heat production is overcome by the stress of excessive cold. Consequently, there are many factors that increase the risk of hypothermia:

\begin{itemize}
    \item Environment
    \begin{itemize}
        \item Cold temperature
        \item Wind
        \item Wet clothing
        \item Altitude
    \end{itemize}
    \item Physiology
    \begin{itemize}
        \item Extremes of age
        \item Poor nutrition
        \item Dehydration
        \item Intoxication
        \item Medications
        \item Immobility
        \item Sleep deprivation
    \end{itemize}
    \item Comorbidities
        \begin{itemize}
            \item Trauma
            \item Hypothyroidism
        \end{itemize}
\end{itemize}

Many activities expose personnel to the cold at Rothera including field work, air travel, boat work, diving, and recreational activities such as skiing. Consequently, all personnel will be exposed to some risk, but the exact degree of risk will vary considerably. Particular high risk groups include divers and boatmen, deep field parties, aircrew and Field Assistants.

Therefore, the level of risk may be described as moderate: whilst the probability of becoming hypothermic is relatively uncommon, the level of harm may vary considerably, and severe harm (i.e. severe hypothermia/death) is rare.

\subsubsection{Risk management}

The most effective strategy is to avoid exposing personnel to the cold. Prescreening of personnel prior to deployment will identify individuals at high risk who may not be suitable for Antarctic service. Other avoidance strategies include provision of adequate shelter/accommodation, provision of well-fitting appropriate clothing and personal protective equipment (PPE), adequate supply of hydration and nutrition, and other equipment specific to the activity (e.g. drysuits or warm-water suits for divers, and snowcats to be preferred over snowmobiles for overland travel in bad conditions).

It is unrealistic to expect personnel to avoid becoming cold at some point whilst deployed in Antarctica, hence mitigation strategies should also be employed to prevent coldness progressing to hypothermia. Hot drinks should be readily available, both on station and in the field, as should calorific snacks. Field parties should be equipped with means of establishing shelter quickly such as a bothy bag or pyramid tent. First aid kits should contain heat-reflective blankets and survival bags to facilitate rewarming. Boating and diving operations should plan to take spare warm clothing in case of becoming wet.

Coupled to both of these elements is education and training, as many of the steps outlined above rely on strict discipline and behaviour modification. Personnel are extensively briefed prior to deployment on fieldcraft to prevent hypothermia, how to recognise it in themselves and others, and how to initiate treatment.

\subsubsection{Management}

The diagnosis of hypothermia may be made either clinically or by measurement of core body temperature with a low-reading thermometer. The Swiss staging system for hypothermia provides the most pragmatic system for diagnosis and may also guide management (Table~\ref{tab:swiss}).\cite{Durrer:2003it} Severe hypothermia (HT IV) is notoriously difficult to distinguish from clinical death, giving rise to the adage ``nobody is dead until they are warm and dead.'' However, in the context of a non-permissive environment where other personnel may be at risk from exposure, a useful discriminator is an incompressible chest wall, indicating that the body is frozen solid and resuscitative efforts are futile.\cite{Brown:2012ja}

A variety of techniques are available for rewarming hypothermic patients:\cite{Brown:2012ja}

\begin{itemize}
    \item Passive rewarming
    \begin{itemize}
        \item Warm environment \& clothes
        \item Warm drinks
        \item Active movement
    \end{itemize}
    \item Active external rewarming
    \begin{itemize}
        \item Heat packs (charcoal/electric)
        \item Forced-air blankets (e.g. Barehugger)
    \end{itemize}
    \item Minimally invasive rewarming
    \begin{itemize}
        \item Warmed (40ºC) intravenous fluids
    \end{itemize}
    \item Invasive rewarming
    \begin{itemize}
        \item Without cardiac support
        \begin{itemize}
            \item Peritoneal dialysis
            \item Haemodialysis
            \item Thoracic lavage
        \end{itemize}
        \item With cardiac support
        \begin{itemize}
            \item Extracorporeal membrane oxygenation (ECMO)
            \item Cardiopulmonary bypass (CPB)
        \end{itemize}
    \end{itemize}
\end{itemize}

\begin{table}
    \begin{tabu} to\linewidth{X[l] X[2,l] X[c] X[2,l]}
        \toprule
        \rowfont\bfseries Stage & Symptoms & Core Temp. (ºC) & Treatment \\
        \midrule
        HT I & Conscious, shivering & 32--35 & Warm environment and clothing, warm sweet drinks, and active movement \\
        HT II & Impaired consciousness, no shivering & 28--32 & Cardiac monitoring, horizontal position \& immobilisation, active external rewarming (heat packs and blankets) \& minimally invasive rewarming (warm IV fluids) \\
        HT III & Unconscious, no shivering, vital signs present & 24--28 & As for HT II + airway management, consider rewarming with ECMO or cardio-pulmonary bypass \\
        HT IV & No vital signs, pliable chest wall & <24 & HT II \& III management + ALS, rewarming with ECMO or cardio-pulmonary bypass \\
        \bottomrule
    \end{tabu}
    \caption{The Swiss staging system for hypothermia\cite{Durrer:2003it}} \label{tab:swiss}
\end{table}

As shown in Table~\ref{tab:swiss}, treatment of hypothermia is guided by the Swiss stage. Nevertheless, management may be broadly divided into prehospital and hospital stages.

In the field, priorities should focus on provision of basic or advanced life support (BLS/ALS), passive and active rewarming, careful handling due to potential cardiac instability, and expedited transfer to the nearest medical facility.\cite{Brown:2012ja} CPR should only be commenced after careful checking for signs of life for at least one minute, and with the understanding that a prolonged resuscitation attempt may be required, potentially lasting hours.\cite{Soar:2010kd} Rescuers should be mindful of this requirement when undertaking such an attempt in a non- or semi-permissive environment. Rewarming in the prehospital setting should focus on heat packs and forced-air blankets, as other methods are either impractical or do not confer substantial heat transfer.\cite{Lundgren:2011ce,Hultzer:2005vn}

In the hospital setting, a hypothermic patient who is haemodynamically stable and normal should undergo active external and minimally invasive rewarming, as described above. This should produce a rise in core body temperature of \~2ºC/hr and be sufficient treatment for stage HT I and II hypothermia.\cite{Auerbach:2012tq,vanderPloeg:2010im} Patients presenting in stage HT III or IV hypothermia will often be haemodynamically compromised and may require cardiac support in the form of ECMO or CPB in addition to the rewarming techniques described for stages HT I and II.\cite{Brown:2012ja}

Hypothermic patients often have large intravenous fluid requirements due to cold diuresis and vasodilation associated with rewarming.\cite{Soar:2010kd} Balanced salt solutions should be used to avoid compounding a pre-existing metabolic acidosis.\cite{PowellTuck:2011us}

\subsubsection{Limitations of standard management}

Treatment of hypothermia in the field is limited to passive and active external rewarming, as invasive rewarming is clearly inappropriate and minimally invasive rewarming requires a prodigious amount of IV fluids. Furthermore, prehospital management may be complicated by difficulties in extricating and transporting a hypothermic patient, which in profound hypothermia (HT III and IV) may trigger cardiac arrest. On station, in addition to measures available in the field, minimally invasive rewarming may also be commenced. There is no capability to provide cardiac support, and over invasive rewarming techniques would be \emph{ad hoc} at best.

Therefore, hypothermic cardiac arrest in Antarctica is probably unsurvivable, owing to difficulties in transferring the patient to station, the lack of expertise and provision of ECMO or CPB, and the inability to provide Level 3 care. Moreover, attempts at resuscitation in the field are likely to place rescuers at risk themselves, despite their natural inclination to try and render first aid.

\subsubsection{Control measures and plan}

Given the limitations discussed above, the following control measures should be implemented:

\begin{boldenumerate}
    \item \textbf{Predeployment screening} \\
    Infirm individuals who are malnourished (e.g. annorexia nervosa) or who have other medical comorbidities increasing their risk of hypothermia should not be cleared for Antarctic service.
    \item \textbf{Predeployment briefing} \\
    The risk of hypothermia should be communicated to personnel and they should be briefed on prevention, recognition, and treatment of hypothermia. The effective ceiling of care should also be made clear.
    \item \textbf{Hypothermic risk assessment for activities} \\
    Comprehensive control measures for specific activities (e.g. diving, air travel etc.) is beyond the scope of this plan. Responsible individuals for these activities should conduct a risk assessment incorporating appropriate measures (e.g. Field Assistants should ensure appropriate shelter is available when planning field trips).
    \item \textbf{Equipment for field medical kits} \\
    Each field medical kit should contain the following items:
    \begin{itemize}
        \item 1 x  Blizzard Rescue Blanket
        \item 1 x North American Rescue Hypothermia Prevention and Management Kit
        \item 1 x TechTrade Ready-Heat 6 Panel Blanket
        \item 5 x 3M Tempadots Disposable Thermometers
        \item 1 x LESS Thermal Hood
        \item 1 x Lifesystems Bothy (4-6 man)
    \end{itemize}
    \item \textbf{Equipment for Rothera station} \\
    For rewarming patients, the medical bay at Rothera should be provisioned with the following items:
    \begin{itemize}
        \item 1 x  low-reading rectal thermometer
        \item 1 x 3M Bair Hugger 775 Temperature Management Unit
        \item 5 x 3M Bair Hugger 300 Full Body Blankets
        \item 1 x 3M Ranger Fluid Warming Unit
        \item 10 x 3M Ranger High Flow Disposable Fluid Warming Sets (Model 24350)
    \end{itemize}
    For prevention or ongoing treatment of hypothermia during medivac, Rothera should also be provisioned with:
    \begin{itemize}
        \item 2 x North American Rescue Hypothermia Prevention and Management Kit
        \item 4 x TechTrade Ready-Heat 6 Panel Blanket
        \item 2 x LESS Thermal Hood
        \item 20 x 3M Tempadots Disposable Thermometers
    \end{itemize}
\end{boldenumerate}

\subsection{Photokeratitis}

\subsubsection{Rationale}

Photokeratitis, colloquially known as snow blindness or arc eye, is an acute inflammatory condition of the sclera following ultraviolet irradiation of the eyes. Whilst it is a self-limiting condition, it is intensely painful -- anecdotally the severity is reported to on par with ureteric colic -- and may be incapacitating for individuals. Unwelcome sequelae include bacterial superinfection with the risk of corneal ulceration, and pterygium -- a benign growth of fibrovascular tissue that may threaten vision.

It is a problem endemic to the polar regions and frequently hampered expeditions during the Heroic Age of Antarctic exploration.\cite{Guly:2012kl} Whilst its incidence is now lower due to education and appropriate precautions, it still affects the unwary or those with inappropriate or ill-fitting eyewear.\cite{BritishAntarcticSurveyMedicalUnit:2013vj}

\subsubsection{Risk assessment}

Photokeratitis is caused by radiation in the ultraviolet (UV) spectrum. Whilst the cornea is transparent to visible light, it absorbs much of the radiation in the UV spectrum, causing epithelial cell death and desquamation. In particular, there appears to be a correlation with exposure to UVB (290-320nm) and, in particular, UVC (<290nm) which is almost entirely absorbed by the corneal epithelium.

Exposure to UV radiation is high in the Antarctic environment, principally due to the large amount of reflected light from ice and snow, and the ozone layer depletion found over Antarctica.\cite{Farman:1985uf} In addition to solar sources, other sources of UV radiation include welding, laboratory UV lamps, and lightboxes used for Seasonal Affective Disorder.

All personnel deployed to Antarctica will be exposed to UV radiation. UV exposure has significant seasonal variation, and personnel deploying for just the Austral winter will have less exposure than personnel during the summer.\cite{Puskeppeleit:1992uv}

Individuals who are at increased risk include those spending large amounts of time outdoors (e.g. Field Assistances, deep field parties),\cite{Moehrle:2003cn} aircrew, staff involved in welding (e.g. mechanical engineers), and personnel using unsafe lightboxes for seasonal affective disorder.

Therefore, the level of risk may be described as moderate: whilst a degree of photokeratitis is not uncommon, the level of harm is usually low (unless it incapacitates an individual to the extent that they place themselves or others at risk).

\subsubsection{Risk management}

As Antarctica is a high UV exposure environment, it is impossible to avoid UV exposure when outside. Consequently all personnel will be exposed to this risk. Avoidance of non-solar UV radiation is more plausible: welding should be kept to a minimum, and BASMU should supply approved lightboxes, removing the need for individuals use their own possibly unsafe devices.

Risk management should therefore focus on mitigating the effects of UV radiation. All personnel should be issued with sunglasses conforming to European Standard EN 1836:2005 with a transmittance rating of 7, indicating full UV protection. Sunglasses should be well-fitting, and either of a wrap-around or leather blinder design to eliminate reflected light from snow or ice. Individuals who who wear contact lenses or glasses should be offered a prescription pair of sunglasses. Personnel who intend to ski or snowboard recreationally should also be provided with advice on obtaining goggles that also conform to this standard.

Personnel involved in welding should be supplied with eye protection conforming to European Standard EN 175:1997.

Personnel should be briefed prior to deployment on the risks of UV radiation, and Field Assistants and party leaders should encourage good eye protection discipline when outside.

\subsubsection{Management}

Photokeratitis is self-limiting, with the corneal epithelium normally regenerating within 72 hours. Management is therefore supportive and should prevent any sequelae.

The diagnosis should be confirmed through history and clinical examination. Important features are foreign body sensation, photophobia, bilateral involvement, and usual antecedent UV exposure due to absence of eye protection. Visual acuity should not be significantly impaired. Inspection usually demonstrates generalised chemosis with no discharge, and the cornea may appear hazy with a miotic pupil. The remainder of the examination should be unremarkable.\cite{Auerbach:2012tq}

Treatment should consist of the following measures:\cite{Auerbach:2012tq}

\begin{itemize}
    \item Analgesia
    \begin{itemize}
        \item Oral analgesia (e.g. Cocodamol and Ibuprufen)
        \item Stat dose topical anaesthetic drops (e.g. Tetracaine)
        \item Eye patching
    \end{itemize}
    \item Reduce photophobia
    \begin{itemize}
        \item Cycloplegic drops (e.g. Cyclopentalate)
    \end{itemize}
    \item Prevent bacterial superinfection
    \begin{itemize}
        \item Lubricant antibiotic ointment (e.g. Chloramphenicol)
    \end{itemize}
\end{itemize}

Topical anaesthetic drops should be used with caution, and only to facilitate initial examination of the eye as a temporising measure until analgesia through other methods is achieved. Prolonged use of anaesthetic drops is contraindicated as it may lead to complications such as corneal ulceration.\cite{Patel:2013ki}

\subsubsection{Limitations of standard management}

All of the treatments described above and diagnostic equipment including an ophthalmoscope and slit lamp are available at Rothera, hence photokeratitis may be managed according to standard techniques when on station.\cite{Marquis:2013vg}

However, photokeratitis often occurs in the deep field, especially after loss or damage to protective eyewear, where the MO may not be on hand to offer a diagnosis. Furthermore, field medical kits only contain chloramphenicol eye ointment and simple analgesia, and are not stocked with other ophthalmic supplies.\cite{Marquis:2013vg}

In the event of loss or damage to sunglasses in the field, eye protection may be improvised by taking a strip of fabric, fashioning it into a blindfold and cutting two slits for the eyes.\cite{Wiseman:2003wx} Deep field parties should take a spare pair of sunglasses as a contingency plan. Similarly, eye patching may be improvised, but symptomatic relief must be balanced against the consequences of impaired vision when in the field.

\subsubsection{Control measures and plan}

The following control measures should be undertaken:

\begin{boldenumerate}
    \item \textbf{Predeployment briefing} \\
    The risk of photokeratitis should be communicated to personnel and they should be briefed on its prevention, recognition, and treatment. The importance of protective eyewear should be stressed.
    \item \textbf{Personal kit}
    \begin{itemize}
        \item Personnel should be issued with well-fitting sunglasses conforming to European Standard EN 1836:2005 with a transmittance rating of 7. They should be of wrap-around or leather blinder style e.g. Cébé Cecchinel sunglasses with Category 4 lenses.
        \item Personnel involved in welding should be with eye protection conforming to European Standard EN 175:1997 e.g. 3M Speedglas 9100 Welding Shield.
    \end{itemize}
    \item \textbf{Equipment for field medical kits}
    \begin{itemize}
        \item 5 x 4g Chloramphenicol eye ointment
        \item 2 x Dressing No. 16 eye pad dressing
        \item 1 x Cébé Cecchinel sunglasses with Category 4 lenses
    \end{itemize}
    \item \textbf{Equipment for Rothera station}
    \begin{itemize}
        \item 20 x 0.5ml 1\% Cyclopentalate eye drops
        \item 20 x 4g Chloramphenicol eye ointment
        \item 20 x 0.5ml 1\% Tetracaine eye drops
        \item 20 x 0.5ml 1\% Fluorescein eye drops
        \item 10 x Dressing No. 16 eye pad dressing
        \item 5 x Cébé Cecchinel sunglasses with Category 4 lenses
        \item 1 x Keeler SL-16 Slit Lamp
        \item 2 x Welch Allyn 97200-BI Elite LED ophthalmoscopes
        \item 2 x Philips goLITE BLU HR3330 lightboxes
    \end{itemize}
\end{boldenumerate}

\subsection{Carbon monoxide poisoning}

\subsubsection{Rationale}

Carbon monoxide is a toxic, invisible, odourless gas that is produced through partial oxidation of carbon compounds; most commonly through the incomplete combustion of fuel due to insufficient oxygen. Carbon monoxide readily binds to haemoglobin to form carboxyhaemoglobin with an affinity \~230 times greater than the oxygen binding affinity.\cite{Townsend:2002cy} Toxicity is primarily mediated by formation of carboxyhaemoglobin, reducing the availability of oxyhaemoglobin leading to hypoxia and death.

Carbon monoxide has a been a persistent hazard to polar exploration, with both Amundsen and Byrd describing narrow escapes from asphyxiation.\cite{Pugh:1959uo} It remains common BAS practice to use traditional Primus stoves for cooking and heating inside pyramid tents in Antarctica, where ventilation may be a secondary consideration in the interests of preserving heat.

Owing to its insidious onset, carbon monoxide poisoning presents a potentially life-threatening risk to Antarctic field operations, as evidenced by previous reports of suspected poisoning at Rothera.\cite{BritishAntarcticSurveyMedicalUnit:2013vj}

\subsubsection{Risk assessment}

Whilst any stove, boiler or engine may produce carbon monoxide, static appliances at Rothera are well ventilated, regularly serviced, and are fitted with carbon monoxide alarms that undergo routine testing. The risk presented by these devices is low.

The principle risk remains the use of stoves within tents, as ventilation within tents is notoriously poor and sometimes purposefully obstructed by occupants in order to preserve heat. Moreover, any carbon monoxide produced is likely to be trapped within the tent where concentrations can easily become toxic.\cite{Thomassen:2004ux,LeighSmith:2005vj,Hampson:2005vj,Betten:2013he} Personnel involved in field operations who spend a protracted time dwelling in tents are exposed to the greatest risk.

One review identified that the risk in tents is further increased with low temperature, altitude, use of large diameter pans, and low flame settings (e.g. simmering).\cite{LeighSmith:2004wr}

Therefore, the level of risk may be described as high: stove use in tents is common, and carbon monoxide poisoning may occur insidiously with all occupants affected. The consequences of unrecognised carbon monoxide poisoning may be catastrophic with multiple fatalities.

\subsubsection{Risk management}

Avoiding the risk is impractical as both tents and stoves are necessary for Antarctic fieldwork. Risk management should therefore focus on mitigating the production of carbon monoxide.

Tent design has an influence on the production of carbon monoxide. The Scott pyramid tent has a breathable inner and outer, along with two ventilation tubes. Additionally, the pyramid structure permits a large temperature gradient between floor and ceiling, contributing to air flow. These features help minimise the production of carbon monoxide.\cite{Betten:2013he,Anonymous:ymdxPOzf}

The choice of fuel also influences carbon monoxide production, as one study suggests heavier fuels increase production.\cite{Betten:2013he} BAS currently uses paraffin which is comparatively heavy and therefore possibly more prone to incomplete combustion. However, it has a large energy density and is safer in terms of flammability than alternatives such as white gas or petrol.

The fitting carbon monoxide detectors to tents should also be undertaken to alert occupants to the presence of carbon monoxide so that they may take appropriate action.

Finally, good campcraft and discipline cannot be overemphasised. Anecdotally, most cases of poisoning resulting from stove use in tents occurs due to human error in blocking off ventilation and not acting on warning signs.\cite{LeighSmith:2004wr} Predeployment training should emphasise the importance of this.

\subsubsection{Management}

The most common signs of carbon monoxide poisoning are headache, nausea and vomiting, vertigo, reduced conscious level, and subjective weakness.\cite{Anonymous:0VMEfFzn} They are non-specific and often insidious in onset, making clinical diagnosis difficult. Definitive diagnosis requires measurement of carboxyhaemoglobin with an arterial blood gas machine.

Immediate treatment of carbon monoxide poisoning is to remove the patient from the source of carbon monoxide and administer high-flow oxygen through a non-rebreathe mask with a reservoir bag to reduce the half-life of carboxyhaemoglobin.\cite{Anonymous:0VMEfFzn} Further management is supportive: patients may be hypotensive requiring IV fluids and vasopressor support, and may also have significant acid-base derangements requiring correction.

Whilst the National Poisons Information Service (NPIS) does not currently recommend routine use of hyperbaric oxygen for carbon monoxide poisoning, this is based on the risk of transferring an unwell patient from one facility to another.\cite{Anonymous:0VMEfFzn} Additionally, a recent Cochrane review concluded that existing RCTs do not establish whether hyperbaric oxygen improves neurological outcome.\cite{Buckley:2011fe} Hence, the role of hyperbaric oxygen remains contentious and requires expert advice.

\subsubsection{Limitations of standard management}

Initial recognition may be delayed due to the non-specific symptoms and the possibility that all people in a tent might be similarly affected. Furthermore, diagnosis would be clinical in the field, as an assay of carboxyhaemoglobin is only available at Rothera.

Initial treatment may also be difficult as inclement weather may mandate the use of shelter and a stove for warmth, making removal of the source of carbon monoxide impractical. Oxygen is not carried by field parties, nor is it available in field medical kits.\cite{Marquis:2013vg} Consequently, oxygen would need to be delivered to field parties first, thereby delaying treatment.

However, once at Rothera poisoning may be managed to a standard approaching that of in the UK. Rothera also has the capability to deliver hyperbaric oxygen, although this would require expert advice from BASMU.

\subsubsection{Control measures and plan}

The following control measures should be adopted:

\begin{enumerate}
    \item Personnel should be briefed on the risks of carbon monoxide, and instructed in proper tent and stove discipline.
    \item Low flame settings and large pans should be avoided when cooking.
    \item Carbon monoxide detectors should be fitted to all tents and Field Assistants should test them weekly.
    \item Stoves should be serviced on a monthly basis.
    \item Daily radio contact should be made with all field parties with a high index of suspicion for symptoms of poisoning.
\end{enumerate}

\pagebreak

\section{Conclusion}

\begin{quote}
Better a live donkey than a dead lion. \\
\em Ernest Shackleton
\end{quote}

As a wilderness environment, Antarctica has many hazards that may challenge the unwary or ill-prepared. This medical plan has outlined a process for rigorously assessing these hazards, undertaken a risk assessment for Rothera station, comprehensively examined three exemplar risks, and presents pragmatic strategies for managing them.

A recurring theme in these control measures is informing personnel of risks that an Antarctic deployment entails, and how they can personally best control their risk. It is impossible for a plan to consider every contingency, but if all team members are appraised of the hazards they face, their joint initiative may facilitate an extemporaneous solution.

\input{appendix.tex}

\bibliographystyle{vancouver}
\bibliography{risk-assessment}

\end{document}
